### Hello!

We're very much looking forward to our 3 days of **R**'ing with you (March 26th - 28th).

In preparation, please install [R](https://rstudio-education.github.io/hopr/starting.html) and [RStudio](https://posit.co/download/rstudio-desktop/). See @sec-install-r for instructions.

Once you have RStudio up and running, please install the following **R packages**. You can just copy and paste the code below into the console of RStudio and hit enter to install.

#### **Day 1 - Data wrangling and visualisation in R**

```{r eval=FALSE}
install.packages(c("tidyverse", "here", "lubridate", "lme4", "ggplot2", "patchwork"))
```

#### **Day 2 & 3 - Statistical Modelling in R**

```{r eval=FALSE}
install.packages(c('ggplot2', 'patchwork', 'DHARMa', 'MASS', 'Hmsc', 'sf', 'terra', 'tidyr', 'vegan', 'dplyr', 'ggcorrplot', 'fields'))
```

```{r eval=FALSE}
remotes::install_version("tmap", "3.3-4")
```

You're also welcome to peruse the course notes linked to in the tabs above, but note that these are subject to change up until day of the course.

Cheers,

Christina, Caitie and Jackson

#### Installing R and RStudio {#sec-install-r}

##### Windows

-   Download R from the [CRAN website](https://cran.r-project.org/bin/windows/base/release.htm).

-   Run the `.exe` file that was just downloaded

-   Go to the [RStudio download page](https://www.rstudio.com/products/rstudio/download/#download)

-   Under *All Installers*, download the RStudio Installer for Windows.

-   Double click the file to install it

-   Once it's installed, open RStudio to make sure it works and you don't get any error messages.

##### MacOS

-   Download R from the [CRAN website](https://cran.r-project.org/bin/macosx/).

-   Select the `.pkg` file for the latest R version

-   Double click on the downloaded file to install R

-   It is also a good idea to install [XQuartz](https://www.xquartz.org/) (needed by some packages)

-   Go to the [RStudio download page](https://www.rstudio.com/products/rstudio/download/#download)

-   Under *All Installers*, download the RStudio Installer for MacOS.

-   Double click the file to install RStudio

-   Once it's installed, open RStudio to make sure it works and you don't get any error messages.

##### Linux

-   Follow the instructions for your distribution from [CRAN](https://cloud.r-project.org/bin/linux), they provide information to get the most recent version of R for common distributions. For most distributions, you could use your package manager (e.g., for Debian/Ubuntu run `sudo apt-get install r-base`, and for Fedora `sudo yum install R`), but we don't recommend this approach as the versions provided by this are usually out of date. In any case, make sure you have at least R 3.3.1.

-   Go to the [RStudio download page](https://www.rstudio.com/products/rstudio/download/#download)

-   Under *All Installers*, select the version that matches your distribution and install it with your preferred method (e.g., with Debian/Ubuntu `sudo dpkg -i rstudio-YYYY.MM.X-ZZZ-amd64.deb` at the terminal).

-   Once it's installed, open RStudio to make sure it works and you don't get any error messages.

#### Updating R and RStudio

If you already have R and RStudio installed, first check if your R version is up to date:

-   When you open RStudio your R version will be printed in the console on the bottom left. Alternatively, you can type `sessionInfo()` into the console. If your R version is 4.0.0 or later, you don't need to update R for this lesson. If your version of R is older than that, download and install the latest version of R from the R project website [for Windows](https://cran.r-project.org/bin/windows/base/), [for MacOS](https://cran.r-project.org/bin/macosx/), or [for Linux](https://cran.r-project.org/bin/linux/)

-   It is not necessary to remove old versions of R from your system, but if you wish to do so you can check [How do I uninstall R?](https://cran.r-project.org/bin/windows/base/rw-FAQ.html#How-do-I-UNinstall-R_003f)

-   Note: The changes introduced by new R versions are usually backwards-compatible. That is, your old code should still work after updating your R version. However, if breaking changes happen, it is useful to know that you can have multiple versions of R installed in parallel and that you can switch between them in RStudio by going to `Tools > Global Options > General > Basic`.

-   After installing a new version of R, you will have to reinstall all your packages with the new version. For Windows, there is a package called `installr` that can help you with upgrading your R version and migrate your package library.

To update RStudio to the latest version, open RStudio and click on `Help > Check for Updates`. If a new version is available follow the instruction on screen. By default, RStudio will also automatically notify you of new versions every once in a while.
